---
title: 10 things I love about Julia
author: Trang Le
date: '2021-12-28'
slug: 10-things-i-love-about-julia
categories:
  - science
tags:
  - julia
  - advent-of-code
  - data science
  - algorithm
description:
  Things I learned and love about Julia through this year's Advent of Code puzzles.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Those who know me know that my love for the [Julia language](https://julialang.org/) has grown quite a bit in the past couple years.
Still, I had trouble finding a project that would allow me to work more in this nifty language
until I saw [Jasmine Hughes](https://twitter.com/Jas_Hughes) solve all of last year [Advent of Code](https://adventofcode.com/) (AOC) puzzles in Julia.
{{%mn%}}I know of Jasmine through the AOC RLadies leaderboard! See Jasmine's valuable advice on "how to get the most out of it" [here](https://youtu.be/336gAu7PAlI).{{%/mn%}}

So that's what I did this year.
Locating in the west coast this time, I have the advantage of not having to stay up until midnight 
or to set the alarm at [4:38 am](https://twitter.com/ClareHorscroft/status/1475469403537952774) 
to be ready when the puzzles come out, which allowed me to end up in the top 3 of the leaderboards I joined. :star: :star: 
So I thought I'd share with you 10 things I love/learned about Julia through AOC.
{{%sn%}}And also some things I find not entirely straightforward.{{%/sn%}}


I hope this post will inspire you to learn Julia.
I won't go in depth into why Julia is awesome
{{%sn%}}For details strong typing, multiple dispatch, environment management and other clever design decisions, please check out this [terrific talk](https://youtu.be/bI9jN40JMQU) by Tamas Nagy instead.{{%/sn%}}
but instead will make some connections to R and thus hopefully eliminate some of the pain points for you if you're learning the language with an R background.
 
_Disclaimers_: I am still very much a novice in Julia and never really had formal programming training.
I did my best to use the right terminology, 
but please let me know if there is a better explanation/description for something.
{{%mn%}} _e.g._, I used _"array"_ and _"collection"_ somewhat interchangibly.{{%/mn%}}
Any new tips/tricks/intuition are most welcomed!
I'd love to update the post with your suggestions!

_Acknowledgements_: 
This post was written in an _.Rmd_ notebook using the wonderful [JuliaCall](https://non-contradiction.github.io/JuliaCall/index.html) package.

## 1. Great packages

Julia has a number of great packages despite being a relatively young language,
much of which can largely be attributed to the [big overlap](https://youtu.be/bI9jN40JMQU) between its package users and developers.
([No two-language problem!](https://juliadatascience.io/julia_accomplish#sec:two_language))

Here are some packages that I frequently call for AOC puzzles:

```{julia, eval=FALSE}
using Chain
using Combinatorics: permutations, combinations
using LinearAlgebra: dot, transpose
using Primes: factor
using StatsBase: countmap, rle
using Statistics
using DataFrames, DataFramesMeta
```

Julia native pipe `|>` supports simple "chaining" operations of one argument,
but I like using the **Chain** package for more comprehensive support with broadcasting and splatting.
For example, for day 15 part 2, I built the larger matrix by

```{julia, eval=FALSE}
append_input(x) = mod1.(input .+ x, 9)
input2 = @chain append_input begin
    map.([i:i+4 for i = 0:4])
    vcat.(_...)
    hcat(_...)
end
```

Among the other functions, I used `countmap()` (≈ R's `table()`) in at least seven of these puzzles.

I leave **DataFrames** and **DataFramesMeta** here in case you're comfortable with wrangling dataframes.
My experience with their syntax so far has not been as fluid as I'd like, so I actually did not use them for any of this year puzzles.
{{%mn%}}This reminded me that `dplyr` is so magical!{{%/mn%}}

## 2. Broadcasting

While R performs "element-wise" operations automatically
{{%sn%}}such as adding a number to a numeric vector, or even adding numeric vectors of different sizes{{%/sn%}}
, in Julia, we have to specify that we want to broadcast by adding a `.` after a function name or before a symbol, such as `.+` or `.==`.

This leads to some cool notations.
For example, if _x_ is a 2-D array and we want to add 1 to all element of _x_:

```{julia, eval=FALSE}
x .+= 1
```

If we have two arrays as arguments for a function, we may need to use `Ref` to ["treat the referenced values as a scalar"](https://discourse.julialang.org/t/how-to-broadcast-over-only-certain-function-arguments/19274/2).
For example:

```{julia, eval=TRUE}
getindex.(Ref([0, 4, 3, 5]), [2, 1, 2])
```

Alternatively, with a wonkier syntax:

```{julia, eval=TRUE}
getindex.(([0, 4, 3, 5],), [2, 1, 2])
```

If you're curious about nested broadcasting, last time I searched, the best way to do it would be 
to define a separate function that perform the inner broadcasting, then broadcast it onto the outer iterable, 
or simply use list comprehension.

## 3. Splat operator

Same as R’s ellipses, Julia's "splat" operator `...` can be used in a function definition to indicate that the function accepts an arbitrary number of arguments.
The Splat operator is also equivalent to using Julia's `reduce()` function (≈ R's `do.call()`).
{{%mn%}}While we're here, Julia's `map()` ≈ R's `apply()`.{{%/mn%}}
For example, here, `max()` returns the maximum of the arguments,
so if we want to find the maximum of an array, we'll need to add `...` after the array in the function argument,
or, for this particular case, use the built-in `maximum()` function.

```{julia, eval=FALSE}
max(4, 5, 3)
max([4, 5, 3]...)
reduce(max, [4, 5, 3])
maximum([4, 5, 3])
```

_Quick note_: `maximum()` can also take a function as its first argument.

Similarly, the following commands are equivalent:
```{julia, eval=FALSE}
*("hello", " ", "world")
*(["hello", " ", "world"]...)
join(["hello", " ", "world"])
```

## 4. Unicode support 

I love that Julia let me use subscript and superscripts in my variable names (with `\_` and `\^`), and Unicode math symbols (with, say, `\alpha<tab>`) even as operations:

```{julia} 
4 ∉ [4, 5, 3]
```
or 
```{julia} 
8 ÷ 3
``` 

A few people on [Zulip](https://julialang.zulipchat.com/) used the octopus emoji 🐙 for AOC 2021 day 11 💥, which made me smile.

## 5. Macros

Some of the macros I frequently used are `@show`, `@time`, `@assert`, and for AOC especially, **BenchmarkTools**'s `@btime`.

If you want, you can even [define your own macros](https://docs.julialang.org/en/v1/manual/metaprogramming/)!

## 6. Scope 

Some developers may not like how Julia deals with variable scope, 
but I actually enjoy being explicit about where I want my variables defined.
In general, it’s best practice to indicate whether a variable name is global or local within your loops. For example:

```{julia, eval=FALSE}
s = 0
for i = 1:10 
    global s += 1
end
```

As of Julia 1.5, if we don’t have `global` here, our code would still run in REPL, 
but if we try to include a _.jl_ script with these lines, Julia will give a warning and also error out.

See more about scoping [here](https://docs.julialang.org/en/v1/manual/variables-and-scoping/).

## 7. Find & filter

When I first learned Julia, I was really frustrated when facing this subsetting problem.
Julia's `findall()` is essentially R's `which()` with a slightly longer syntax.
And if you want the _elements_ instead of _positions_, you can use `filter()`.

```{r, eval=FALSE}
# In R
x <- c(2, 4, 5) 
which(x > 2) ## 2 3
x[x > 2] ## 4 5
```


```{julia, eval=FALSE}
# In Julia
x = [2, 4, 5]

findall(xᵢ -> xᵢ > 2, [2, 4, 5]) ## 2 3
findall(>(2), x) ## 2 3

filter(xᵢ -> xᵢ > 2, [2, 4, 5]) ## 4 5
filter(>(2), x) ## 4 5
x[x .> 2] ## 4 5 (the "R" way)
# x[begin:end .∈ Ref(findall(>(2), x))]
# not what I would do unless I have a list of indices I want to include/exclude
```

One potential gotcha to keep in mind is that, if the condition is not satisfied with any element of the collection,
`findfirst()` returns **nothing** while `findall()` returns an **empty collection**.

A related and useful function here is `indexin()`.

`findall()` can also be used on strings with the argument **overlap** crucial for some AOC puzzles.

## 8. Simultaneous assignments

```{julia, eval=FALSE}
ni, nj = size(input)
```
or 
```{julia, eval=FALSE}
[f(x, y) for (x, y) in my_tuples]
```


## 9. Memoization is easy!
```{julia, eval=FALSE}
using Memoize
@memoize function my_recursive_func
...
end
```


## 10. Other neat things

These are some of the short functions in `utils.jl` in my `julia-aoc` repo:

```{julia, eval=FALSE}
chr2hex(x) = string(parse(Int, x, base = 16), base = 2, pad = 4)
make_mat(input) = make_int(hcat(collect.(input)...))
make_int(s) = parse.(Int, s)
bin2dec(x) = parse(Int, x, base = 2)
sort_vals(counts) = sort(collect(counts), by = x -> x[2])
```

Additional, useful functions/notes for AOC

- `mod1()` FTW!
- `pop!()`, `push!()`, `popfirst!()`, `pushfirst!()`, `filter!()`, `delete!()`, `deleteat!()`
- `collect()` turns object into a collection, ≈ Python's `list()`.
- `Base.product()` ≈ R's `expand_grid()`
- `zip()` ≈ Python's `zip()`
- Multiple dispatch (≈ R's S3 and S4 methods)
- `try ... catch; ... end` can be slow. Check with if else if known
- `end` can be used to index into an array, e.g. `x[1:end-1]` (≈ R's `head(x, -1)`)
- `fill(a (x, y, z))` fills an array of dimension _x_, _y_, _z_ with _a_
- `merge()` to merge dictionaries
- `vcat(x, y)` or `[x; y]` ≈ R's `c(x, y)`
- `findmin(skipmissing(x))` to find minimum while ignoring missing values
- Integer array is faster than character array is faster than string arrays
- While most objects in R are immutable, this is not the case in Julia.
Be sure to use `copy()` or `deepcopy()` when you need to create a copy of, say, an array.
- Know your recursion
- There will always be a Dijkstra problem. 
Brush up on your implementation of the algorithm or learn to use one of the packages that do.
In R, **igraph** is a great package for this.
I still need to try out Julia's **Graphs** package that seems to have an extensive list of [shortest path algorithms](https://juliagraphs.org/Graphs.jl/dev/pathing/#Shortest-Path-Algorithms).

## What's next
I'd like to keep learning Julia, of course.
Most immediately, I'd like to learn about
- how to create and use `struct`
- `CartesianIndices`
- when I should use `convert()`, `Int()`, `string()`, etc.
- `SparseArrays` 

Maybe I'll see you at Advent of Code next year! 🌻


